Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-12 paper.tex
chore: leave mean field but move CTDE in background

One interesting application of \ac{RL} is when there are multiple learning agents involved. 
 Such scenarios are referred to as multi-agent reinforcement learning (MARL)~\cite{zhang2019marl}. 
%
In particular, in this work, we consider homogenous \ac{MAARL}~\cite{yang2021many}, 
 where the set of agents $N \gg 2$ and each agent is interchangeable and indistinguishable.
%
This research area is relevant in the context of large-scale systems 
 where collective intelligence emerges from local and repeated interaction of simple entities, like in swarm robotics.

In fact, the first works in this direction were discussed precisely swarm robotics, %\lukas{what does this mean `in the last field'?}
 exploring new models (e.g., swarMDP~\cite{DBLP:conf/atal/SosicKZK17}) and learning algorithms capable of extrapolating a policy representing the entire system.  \lukas{This sentence still needs rework.}
%Modern approaches, however, 
% have started to consider the use of 
 Recently, deep learning approaches have been considered 
 to synthesise robust controllers capable of generalizing to new tasks. 
%In this context, mean-field reinforcement learning~\cite{pmlr-v80-yang18d} is certainly noteworthy. 
 Mean-field RL~\cite{pmlr-v80-yang18d}  is a technique for learning approximate inference in large-scale systems. Here, the interactions between the agents can be decoupled into a single conditional probability distribution function for each agent, 
 thus simplifying the learning process and reducing the computational cost.
%
Some known approaches using mean-field reinforcement learning include Q-mean, 
 which is an extension of Q-learning to mean-field settings~\cite{yang2018mean}, 
 and actor-critic mean-field~\cite{frikha2023actor}, which combines actor-critic algorithms with mean-field approximations. 
%
These approaches have shown promising results in various domains, such as multi-agent coordination 
 and decentralised control, and are actively being researched and developed for further applications.
\subsection{Problem formalisation}
Given the homogeneity, large system scale, and the \emph{locality} (i,e., each agent can only observe its neighbours), 
 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17} -- 
 an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model for swarm-like systems.
% 
A SwarMDP is characterised by a \emph{swarming agent} ($\mathbb{A}$) and the dynamics of the environment ($\mathbb{E}$).
Specifically, $\mathbb{A}$ is a tuple ($\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, \pi$) where:
\begin{itemize}
  \item $\mathcal{S, O, A}$ are the set of local states, observations (or features), and actions, respectively;
  \item $\mathcal{R}: \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, which is influenced by the environment;
  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions. It could be deterministic or stochastic.
\end{itemize}
Starting from this definition, the environment $\mathbb{E}$ is defined as a tuple ($\mathcal{P}, \mathbb{A}, \mathcal{T}, \xi$), where:
\begin{itemize}
  \item $\mathcal{P}$ is the total number of agents in the systems (the agent population), which is assumed to be fixed;
  \item $\mathbb{A}$ is the defined agent prototype that rules each agent $v \in P$;
  \item $\mathcal{T}: \mathcal{S}^P \times \mathcal{A}^P \times \mathcal{S}^P \rightarrow \mathbb{R}$ is the transition  global function, which is influenced by the actions of the agents and returns a collective reward -- this is typically not known by the swarming agents;
  \item $\xi: \mathcal{S^P} \rightarrow \mathcal{O^P}$ is the global observation model of the systems.
\end{itemize}
In swarMDP, the neighbourhood is not directly defined, but it is implicitly defined by the observation model $\xi$.
In our specific case, the agents can only interact with 1-hop neighbourhoods and are not directly influenced by other agent observations. We can therefore restrict the observation model as follows:
\begin{equation}
\xi(v): \{s_j, j \in \mathcal{N}^v\} \rightarrow O
\end{equation}
\begin{equation}
\xi = \{\xi(v), v \in \mathcal{P}\}
\end{equation}
where $\mathcal{N}^v$ is the set of neighbours of $v$.

@@

In swarm-like systems, 
 the typical approach for learning follows a strategy known as \ac{CTDE}. \ga{perhaps it should be introduced in many RL} 

@@

In our case, we leveraged the property of \ac{GNN}s to have a dual interpretation, i.e., to function globally over the entire graph and locally only over the neighbourhood. 
%
Therefore, we relied on the value-based approach \ac{DQN}~\cite{mnih2015playing} \ga{to introduce in many rl??} with two major modifications (see \Cref{alg:dqn}): 
\begin{enumerate}
  \item experience replay stores experiences in the form of \emph{graphs},
  \item the neural network used to compute the Q function is based on a \ac{GNN} with an \ac{MLP} downstream.
\end{enumerate}
The first point is a natural extension because we work on graphs rather than simple values. 
%
For the second point,
 the use of \acp{GNN} allows us to define policies on a variable neighbourhood  that is essential in such systems as this can vary based on the neighbourhood policy,
 and it is known that \acp{GNN} have a certain ability to generalise to new structures and scale with different nodes. 
% 
Additionally, 
 using the overall graph compared to local experiences makes learning more stable as it reduces the non-stationarity of the environment perceived by each node.
% 
 During learning, it is as if we have a \emph{single} agent collectively choosing which action to take (i.e., resulting action graph). 
%
\begin{algorithm}
  \KwIn{Environment $\mathbb{E}$, graph replay buffer $\mathcal{D}$, target network $\theta^{-}$, current network $\theta$, exploration strategy $\epsilon$}
  \KwOut{Trained DQN model $\theta$}

>>>>

One interesting application of \ac{RL} is when there are multiple learning agents involved. 
 Such scenarios are referred to as multi-agent reinforcement learning (MARL)~\cite{zhang2019marl}. 
In particular, in this work, we consider homogenous \ac{MAARL}~\cite{yang2021many}, 
 where the set of agents $N \gg 2$ and each agent is interchangeable and indistinguishable.
This research area is relevant in the context of large-scale systems 
 where collective intelligence emerges from local and repeated interaction of simple entities, like in swarm robotics.
In such many-agent scenarios, 
 the implementation of fully decentralized learning is often unfeasible due to the large number of learning agents, 
 which renders the system non-stationary and difficult to manage. 
Conversely, a centralized controller capable of coordinating 
 the entire system may not be a viable solution due to scalability concerns. 
 To address this challenge, a practical solution is the adoption of \ac{CTDE}.
%In fact, the first works in this direction were discussed precisely swarm robotics, %\lukas{what does this mean `in the last field'?}
% exploring new models (e.g., swarMDP~\cite{DBLP:conf/atal/SosicKZK17}) and learning algorithms capable of extrapolating a policy representing the entire system.  \lukas{This sentence still needs rework.}
%Modern approaches, however, 
% have started to consider the use of 
% Recently, deep learning approaches have been considered 
% to synthesise robust controllers capable of generalizing to new tasks. 
%In this context, mean-field reinforcement learning~\cite{pmlr-v80-yang18d} is certainly noteworthy. 
Mean-field RL~\cite{pmlr-v80-yang18d} is one of such concrete applications of \ac{CTDE} 
 where the interactions among the population of agents are estimated by considering either the effect of a single agent and the average impact of the entire population or the influence of neighbouring agents.
Some known approaches using mean-field reinforcement learning include Q-mean, 
 which is an extension of Q-learning to mean-field settings~\cite{yang2018mean}, 
 and actor-critic mean-field~\cite{frikha2023actor}, which combines actor-critic algorithms with mean-field approximations. 
These approaches have shown promising results in various domains, such as multi-agent coordination 
 and decentralised control, and are actively being researched and developed for further applications.
\subsection{Problem formalisation}
Given the homogeneity, large system scale, and the \emph{locality} (i,e., each agent can only observe its neighbours), 
 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17} -- 
 an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model for swarm-like systems.
A SwarMDP is characterised by a \emph{swarming agent} ($\mathbb{A}$) and the dynamics of the environment ($\mathbb{E}$).
Specifically, $\mathbb{A}$ is a tuple ($\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, \pi$) where:
\begin{itemize}
  \item $\mathcal{S, O, A}$ are the set of local states, observations (or features), and actions, respectively;
  \item $\mathcal{R}: \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, which is influenced by the environment;
  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions. It could be deterministic or stochastic.
\end{itemize}
Starting from this definition, the environment $\mathbb{E}$ is defined as a tuple ($\mathcal{P}, \mathbb{A}, \mathcal{T}, \xi$), where:
\begin{itemize}
  \item $\mathcal{P}$ is the total number of agents in the systems (the agent population), which is assumed to be fixed;
  \item $\mathbb{A}$ is the defined agent prototype that rules each agent $v \in P$;
  \item $\mathcal{T}: \mathcal{S}^P \times \mathcal{A}^P \times \mathcal{S}^P \rightarrow \mathbb{R}$ is the transition  global function, which is influenced by the actions of the agents and returns a collective reward -- this is typically not known by the swarming agents;
  \item $\xi: \mathcal{S^P} \rightarrow \mathcal{O^P}$ is the global observation model of the systems.
\end{itemize}
In swarMDP, the neighbourhood is not directly defined, but it is implicitly defined by the observation model $\xi$.
In our specific case, the agents can only interact with 1-hop neighbourhoods and are not directly influenced by other agent observations. We can therefore restrict the observation model as follows:
\begin{equation}
\xi(v): \{s_j, j \in \mathcal{N}^v\} \rightarrow O
\end{equation}
\begin{equation}
\xi = \{\xi(v), v \in \mathcal{P}\}
\end{equation}
where $\mathcal{N}^v$ is the set of neighbours of $v$.

@@

Since we consider swarm-like systems, that are an example of many-agent systems, 
 the proposed approach follows a \ac{CDTE} learning pattern,
 but differently from mean-field approaches and actor-critic solutions,
 we use a value-based approach combined with a \ac{GNN} as a function approximator.
%In swarm-like systems, 
% the typical approach for learning follows a strategy known as \ac{CTDE}. \ga{perhaps it should be introduced in many RL} 
%The idea is to learn a policy at simulation time where there is a collective view of the system, 
% and then at runtime use the policy found with global information. 
%This approach allows policies that are influenced by global information but only require local information to function at runtime. 

@@

%The typical approach in such cases is based on actor-critic systems~\cite{DBLP:conf/nips/LoweWTHAM17,wu2022more,song2022ctds,song2022centralized},
%  where the \emph{actor} is the distributed policy (with only local information) and the critic is a neural network that takes the overall system state.
Specifically, we leveraged the property of \ac{GNN}s to have a dual interpretation, i.e., to function globally over the entire graph and locally only over the neighbourhood. 
As value-based algorithm we relied on \ac{DQN}~\cite{mnih2015playing} \ga{to introduce in many rl??} with two major modifications (see \Cref{alg:dqn}): 
\begin{enumerate}
  \item experience replay stores experiences in the form of \emph{graphs},
  \item the neural network used to compute the Q function is based on a \ac{GNN} with an \ac{MLP} downstream.
\end{enumerate}
The first point is a natural extension because we work on graphs rather than simple values. 
For the second point,
 the use of \acp{GNN} allows us to define policies on a variable neighbourhood  that is essential in such systems as this can vary based on the neighbourhood policy,
 and it is known that \acp{GNN} have a certain ability to generalise to new structures and scale with different nodes. 
Additionally, 
 using the overall graph compared to local experiences makes learning more stable as it reduces the non-stationarity of the environment perceived by each node.
 During learning, it is as if we have a \emph{single} agent collectively choosing which action to take (i.e., resulting action graph). 
\begin{algorithm}
  \KwIn{Environment $\mathbb{E}$, graph replay buffer $\mathcal{D}$, target network $\theta^{-}$, current network $\theta$, exploration strategy $\epsilon$}
  \KwOut{Trained DQN model $\theta$}
